---
title: "Disparities"
author: "Alison Turner"
date: '2022-05-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Disparities in Sentencing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(readxl)
library(tidyverse)
library(censusapi)
library(lubridate)
library(foreach)
library(tigris)
library(stringr)
library(sf)



# Functions for data processing
# These functions are used to clean ACS pulled data 

yeargeog <- function(year,state,vars){
  data.frame(year = year, 
             getCensus(name="acs/acs5",
                       vintage=year,
                       vars = vars,
                       key=key,
                       region = "tract:*",
                       regionin = paste0("state:",unique(fips_codes$state_code)[state])
             )
  )
  
}

fips_to_acs <- function(data){
  t <- left_join(data,fips_codes,by=c("state"="state_code","county"="county_code"))
  t
}


```


```{r Data Processing}
IDOC <- read_xls("Restore Justice/June 2017 Prison Stock.xls",skip = 5)


IDOC$yeartrans <- ifelse(grepl("^[[:digit:]]+$",IDOC$`Sentence Years`),IDOC$`Sentence Years`,NA)
IDOC$yeartrans <- as.numeric(IDOC$yeartrans)

#2017 requires a dob change in format
IDOC$`Date of Birth` <- mdy(IDOC$`Date of Birth`)
IDOC$`Sentence Date` <- mdy(IDOC$`Sentence Date`)

IDOC <- IDOC %>% mutate(age=year(Sys.Date())-year(`Date of Birth`),
                        life=ifelse(`Sentence Years`=="LIFE",1,(ifelse(yeartrans>49,"De Facto",0))),
                        sdp=ifelse(`Sentence Years`=="SDP",1,0),
                        "Sentence Age"=year(`Sentence Date`)-year(`Date of Birth`)) %>%
  select(-yeartrans)


```


```{r ACS pull}
#replace with a local census key 
key <- "095f223bd2d9dd69b8546222bde6c171c540da9b"

race_vars <- c("B01001A_001E","B01001B_001E","B01001C_001E","B01001D_001E","B01001E_001E","B01001F_001E","B01001G_001E","B01001H_001E","B01001I_001E")

acs5_race_17 <- getCensus(name = "acs/acs5",
                       vintage = 2017,
                       vars = race_vars,
                       key = key,
                       region="county:*",
                       regionin = "state:17")
acs5_race_17$year <- 2017

acs5_race_20 <- getCensus(name = "acs/acs5",
                          vintage = 2020,
                          vars = race_vars,
                          key = key,
                          region="county:*",
                          regionin = "state:17")
acs5_race_20$year <- 2020

acs5_race <- rbind(acs5_race_17,acs5_race_20)

acs5_race <- fips_to_acs(acs5_race)
acs5_race <- acs5_race %>% select (year,state_name,county.y,"B01001A_001E","B01001B_001E","B01001C_001E","B01001D_001E","B01001E_001E","B01001F_001E","B01001G_001E","B01001H_001E","B01001I_001E" )

names(acs5_race) <- c("year","state","county", 
                      "white alone",
                      "black alone",
                      "american indian or alaska native alone",
                      "asian alone",
                      "native hawaiian or pacific islander alone",
                      "some other race alone",
                      "two or more races",
                      "white alone, not hispanic or latino",
                      "hispanic or latino, any race category"
)


#transformation of data frame to align with IDOC data 
acs5_race$county <- str_remove(acs5_race$county," County")
acs5_race$total <- rowSums(acs5_race[,4:10])
acs5_race_perc <- data.frame(acs5_race[,1:3],lapply(acs5_race[,4:10],function(x) (x/acs5_race$total)*100))
acs5_race_perc$check <- rowSums(acs5_race_perc[,4:10])
15
IL_counties <- tigris::counties(state=17) %>%
  select(county = NAME,geometry) %>% 
  st_transform(4326) 

names(acs5_race_perc)[4:10] <- c("White","Black","American Indian","Asian","Native Hawaiian","Unknown","Bi-Racial")


ACS_IL <- acs5_race_perc %>% 
  select(-check) %>% 
  pivot_longer(cols = 4:10,names_to = "Race")


ACS_IL <- left_join(ACS_IL,IL_counties)

```


In Illinois, the extreme sentencing of youth to life and de facto life isn’t just an urban issue. Most counties have dispensed at least one of these extreme sentences to a currently serving individual.

That said, some counties dispense these extreme sentences at rates disproportionate to their share of the population and black youth bear a disproportionately large burden of the state’s youth life and de facto life sentences.



### Points of Note

    In Illinois, twelve counties comprising 74% of the state’s total population are responsible for almost 90% of all life and de facto sentences for offenses before age 26.
    Four counties—Peoria, Macon, Cook, and St. Clair—contribute a disproportionately large share of individuals to the total population of Illinois youth granted extreme sentences, relative to their share of state population.
    In Illinois, 68% of the individuals serving life and de facto life sentences for offenses before age 26 are black.
    For some counties, black youth bear a share of dispensed youth life and de facto life sentences roughly three to seven times greater than the black share of those counties’ total populations.

### Concentrated Sentencing 


For some counties, extreme youth sentences are dispensed at rates higher than the county’s share of the state’s population.



In Illinois, not all counties sentence youth to life or de facto life at rates proportionate to their share of the state’s population.

Rather, the top twelve counties—defined as counties that have sentenced 20 or more currently serving individuals to these extreme terms—are responsible for almost 90% of the entire state’s youth life and de facto life population.

Table 1 compares the population share of each county to its share of individuals currently serving life and de facto life for offenses before age 26.


Of these, four counties—Peoria, Macon, Cook, and St. Clair—contribute a disproportionately large share of individuals to the total pool.

Overall, these twelve counties are diverse. Three counties—DuPage, Will, and Lake—rank among Illinois’s most prosperous, measured by median household income. Five others—St. Clair, Macon, Cook, Winnebago, and Peoria—have some of the state’s highest rates of household poverty. Beyond economics, these counties are also distributed across Illinois, and range dramatically in their politics, demographics, and share of population.

Given these differences, to what degree might disparities reflect a specific and disproportionate harshness towards youth during sentencing, rather than harsher sentencing overall? To move toward an answer, Restore Justice evaluated each county’s extreme youth sentences in the context of its broader sentencing trends.

Table 2 compares each county’s share of Illinois’ extreme youth sentences to their share of life and de facto sentences dispensed to individuals above the age of 26.


These data show that for at least some counties, the disproportionate sentencing of youth to life and de facto life cannot be explained solely by a broader, county-wide trend towards the dispensing of harsher sentences.


## Racial Disparities


Black youth disproportionately bear the burden of Illinois’ life and de facto life sentences.



In Illinois, roughly 68% of the individuals serving life and de facto life sentences for offenses before age 26 are black, even though black Illinoisans comprise just 15% of the state’s total population.

This pattern of racial disparity reiterates at the county level.

All of the twelve Illinois counties that have sentenced more than 20 currently serving individuals to youth life or de facto life dispensed those sentences disproportionately to black youth.

This dynamic is represented in Figure 1, which compares for each county (1) the proportion of that county identified as black, as determined by the U.S. Census, versus (2) the proportion of that county’s extreme youth sentences dispensed to black youth.

#### Figure 1. Black Share of County Population vs. Extreme Sentences to Black Youth at Top Twelve Illinois Counties


If a county’s demographics should roughly reflect that of its incarcerated youth, these twelve counties sentence black youth to life and de facto life at rates between three and seven times in excess of those expectations.
